Do motif roles provide unique information about species risk of extinction?

### 2021-06-04 manuscript/motifs_vs_degree.tex

	\subsection*{Calculating persistences}

		First, we assign all species a baseline probability of extinction ($\pi_{base}$) -- the risk of going extinct given factors not related to the food web itself. 
		For consumer species, this also applies to extinction despite all resources being present. 
		Such events can for example reflect diseases or stochastic extinctions of small populations.

		Additionally, we analyze a set of scenarios where primary producers get a higher baseline probability of extinction ($\pi_{disturbed}$). 
		This reflects a threat to primary producers such as increased temperatures resulting in droughts, making some species more vulnerable to extinction.
		$\pi_{disturbed}$ ranged between $0.1-0.5$, in steps of $0.08$. The highest disturbance level, $\pi_{disturbed} = 0.5$, corresponds to basal species having a 50\% risk of going extinct. 
		Consumer species retained $\pi_{base}=0.1$ in all cases.

@@

        We then describe meso-scale structure at the species level by defining each species' motif participation. [[Should stop calling this motif roles]]
		Finally, we defined species' motif participation following~\citet{Stouffer2012}. 
		For each network, we identified each group of three interacting species and assigned it to the corresponding motif. 
		A species' motif participation is a vector of the number of times it appears in each motif.
		We calculated species' motif participations using the Python package \emph{pymfinder}~\citep{pymfinder}.
        As with motif profiles of networks, we normalized participation vectors by dividing each count by the total number of motifs the species appears in.
        Note, however, that this normalization does \emph{not} control for differences due to degree \emph{per se} as high-degree species also tend to participate in different types of motifs than low-degree species~\citep{Cirtwill2021_inprep}.

		We began by calculating species' in-degrees (numbers of prey).
		As the Bayesian network approach only accounts  bottom-up effects, out-degree (number of predators) is irrelevant to extinction risk.
		Note, however, that out-degree may also be important in topological simulations which include both top-down and bottom-up effects.

        Next, we calculated species' trophic levels. 
		Among the many definitions of trophic level available, we chose shortest trophic level (STL): the length of the shortest food chain between the focal species and any basal resource~\citep{Williams2004}. 
		Basal resources have an STL of one, any species consuming a basal resource has an STL of two, etc.
        STL is therefore a measure of the minimum number of steps away a species is from any disturbance to the base of a food web.

@@

            We are interested in whether having a greater proportion of a species' role be made up of a certain motif (e.g., direct competition) might increase persistence (see \emph{Appendix SXX} for relationships between persistence and counts of motifs).
            To test this, we fit four linear mixed-effect models (LMMs; one per motif included in the Bayesian networks).
            These models included the proportion of the species' role made up by the focal motif, the level of disturbance to basal species, and the interaction between the two.
            We centered and scaled the predictors before fitting the LMMs.
            The models also included a random intercept for the interaction between species richness and connectance (equations.
            All models were fit using the R~\citep{R} function `lmer' from the package \emph{lmerTest}~\citep{lmerTest}.
            We calculated marginal (fixed effects only) and conditional (fixed and random effects) $R^2$ using the R~\citep{R} function `r.squaredGLMM' from the package \emph{MuMIn}~\citep{MuMIn}.

@@

            Without disturbance, a species' persistence increases with increasing proportion of the omnivory and direct competition motifs. Conversely, a species' persistence decreases with increasing proportion of the apparent competition motif. Finally, a species persistance is  negatively but not significantly related to the proportion of the three-species chain (Figure~\ref{fig:prop_lmer_all}).

            At high levels of disturbance, a species' persistence decreases with increasing proportion of the three-species chain, but increases with increasing proportion of all other motifs (Figure~\ref{fig:prop_lmer_all}).
            This means that increased disturbance strengthen the effects of the three-species chain motif and changes direction of the effect from the apparent competition motif  (Table S4, \emph{Appendix SD}).
            The effect of omnivory on persistence is still positive but weaker at higher levels of disturbance, and the effect of direct competition are largely unaffected by disturbance level.

@@

            By fitting linear regressions of persistence against motif proportions for the species for each network separately, we can extract the slopes and analyze their distribution. This reveals information on how the relationship change with the level of disturbance and global network structure. For the global network parameters the network size had a much smaller effect than connectance (Fig.~\ref{fig:density_prop_S}; \emph{Appendix SZ})and therefore we here focus on only connectance.

            Networks with high connectance generally have more consistent relationships between persistence and motif participation, i.e., sharper peaks of the density distributions (Fig.~\ref{fig:density_prop_C}). However, the interaction between a species' participation in the various motifs and disturbance changes with connectance  (Figure~\ref{fig:density_prop_C}).
            This is in particular evident for the direct competition and omnivory motif. At low and medium connectances there is little change in the proportion of positive slopes (\textgreater80\%) for the different levels of disturbance. However, at high connectances the proportion of positive slopes is overall lower and also decreases with increasing disturbance. This means that at low levels of connectance, in a majority of cases an increased proportion of the direct competition motif increase persistence, while the opposite is true with high connectance. 
            The trend is reversed for the omnivory motif.
            Overall, the proportion of networks showing a positive effect of omnivory decreases with increasing disturbance.
            This decrease is strongest at low and moderate connectances but very weak at high connectance. (Figure~\ref{fig:density_prop_C}). Notably, the distribution of slopes for the omnivory motif is wider (effect is more variable) the lower the connectance is.
            Hence, the level of connectance influence whether a high proportion of the omnivory motif is associated with higher persistence or not.

@@

            \begin{table}[hb!]
            \centering
            \caption{The meso-scale composition of a network varies with size and connectance. Here we give the range and mean ($\pm$ standard deviation) proportion of each motif in species' motif participation vectors. 
            We also show the effects ($\beta$) of network size, connectance, and their interaction in linear models predicting the proportion of each motif in a participation vector.
            All effects were significant ($p$\textless0.001).
            Predictors were not centred or standardized before model fitting. Motifs are ordered according to mean proportion.}
            % Relevant lms are pchainlm, pomnilm, papplm, pdirlm}
            \label{tab:partic_vs_SC}   
            \footnotesize
            \begin{tabular}{c|c c c | c c c}
                Motif & Min & Max & Mean (SD) & Size & Connectance & Interaction \\
                \hline
                App. Comp & 0.250 & 0.765 & 0.422 (0.0645) & -5.06$\times10^{-4}$ & -1.00 & 2.59$\times10^{-3}$ \\
                Chain & 0.109 & 0.413 & 0.235 (0.042) & 1.78$\times10^{-4}$ & -0.157 & -3.27$\times10^{-3}$ \\
                Dir. Comp & 0.0501 & 0.359 & 0.179 (0.0407) & -3.56$\times10^{-5}$ & -0.556 & 5.92$\times10^{-3}$ \\
                Omnivory & 0.006 & 0.346 & 0.164 (0.0790) & 3.64$\times10^{-4}$ & 1.71 & -5.24$\times10^{-3}$ \\   
                \hline
                \end{tabular}
                \end{table}

        The numbers of different motifs in a network may vary with web size, connectance, and their interaction.
        To establish this context, we fit linear models relating the proportion of each motif in a species' role against network size, connectance, and their interaction.

        \subsubsection*{Motif participation and simple role measures}

            Species' motif roles also vary with their in-degree and trophic level (Fig.~\ref{fig:motifs_vs_TL_and_deg}).
            Species with higher in-degree tend to have higher proportions of their motif role made up by omnivory. Proportions of the other three motifs tend to decline with increasing degree.

@@

        \begin{figure}
            \centering
            \includegraphics[width=.95\textwidth]{figures/persistence_positions_byprey.eps}
            \caption{A species' probability of persistence varied with the frequency of different motif positions in its role. Relationships are shown for the range 0-0.8 for all positions, but some positions (especially in the omnivory motif) are less common. Line colour indicates the motif containing a position. Line style indicates the number of prey within the motif for each position. Symbol shape indicates the number of predators within the motif for each position; note that only bottom-up extinctions were included in our simulations. The top set of lines corresponds to a probability of extinction for basal resources of 0.1. The middle set of lines correspons to a probability of extinction for basal resources of 0.26, and the lowest set of lines corresponds to a probability of extinction for basal resources of 0.5.}
            \label{fig:persistence_positions_prey}
        \end{figure}

@@

        [[have fit these models but no figure yet - seems like degree and the 2-prey positions show the same pattern so they're probably tightly linked. STL may explain the differences in top and middle positions for the chain. There's still a fair amount of differences among the 0 prey and 1 prey motifs though, so motifs still provide unique information that degree and STL don't.]]

>>>>

	\subsection*{Calculating persistence}

		First, we assigned all species a baseline probability of extinction ($\pi_{base}$) -- the risk of going extinct given factors not related to the food web itself. 
		For consumer species, this also applies to extinction despite all resources being present. 
		Such events can, for example, reflect diseases or stochastic extinctions of small populations.

		As well as this baseline scenario, we simulate a set of scenarios where primary producers get a higher baseline probability of extinction ($\pi_{disturbed}$). 
		These scenarios reflect threats to primary producers such as increased temperatures resulting in droughts, making some species more vulnerable to extinction.
		$\pi_{disturbed}$ ranged between $0.1-0.5$, in steps of $0.08$. The highest disturbance level, $\pi_{disturbed} = 0.5$, corresponds to basal species having a 50\% risk of going extinct. 
		Consumer species retained $\pi_{base}=0.1$ in all cases.

@@

        We then describe meso-scale structure at the species level by defining each species' motif participation and species' roles.
        Motif participation describes the frequency with which a focal species appears in each of the four motifs, analogous to the network motif profile~\citep{Stouffer2012}.
        A species' role describes the frequency with which a focal species appears in each \emph{unique position} (e.g., top, middle, or bottom species in the three-species chain, top or bottom species in the apparent competition motif) within each motif ~\citep{Stouffer2012,Cirtwill2017}.
		We calculated species' motif participation and roles using the Python package \emph{pymfinder}~\citep{pymfinder}.
        As with motif profiles of networks, we normalized participation (and role) vectors by dividing each count by the total number of motifs (motif positions) the species appears in.
        Note, however, that this normalization does \emph{not} control for differences due to degree \emph{per se} as high-degree species also tend to participate in different types of motifs than low-degree species~\citep{Cirtwill2021_inprep}.

        Finally, we calculated simpler measures of a species' role within its community: degree and trophic level.
		As the Bayesian network approach only accounts  bottom-up effects, out-degree (number of predators) is irrelevant to extinction risk.
		We therefore focused on species' in-degrees (number of prey).
		Note, however, that out-degree may also be important in topological simulations which include both top-down and bottom-up effects.

		There are many definitions of trophic level available, most of which tend to be very similar~\citep{Carscallen2012}.
		Here we use shortest trophic level (STL): the length of the shortest food chain between the focal species and any basal resource~\citep{Williams2004}. 
		Basal resources have an STL of one, any species consuming a basal resource has an STL of two, etc.
        STL is therefore a measure of the minimum number of steps away a species is from any disturbance to the base of a food web.

@@

            We are interested in whether having a greater proportion of a species' role be made up of a certain motif (e.g., direct competition) might increase persistence.
            % (see \emph{Appendix SXX} for relationships between persistence and counts of motifs).
            To test this, we fit four linear mixed-effect models (LMMs; one per motif included in the Bayesian networks).
            These models included the proportion of the species' role made up by the focal motif, the level of disturbance to basal species, and the interaction between the two.
            We centered and scaled the predictors before fitting the LMMs.
            The models also included a random intercept for the interaction between species richness and connectance (equations.
            All models were fit using the R~\citep{R} function `lmer' from the package \emph{lmerTest}~\citep{lmerTest}.
            We calculated marginal (fixed effects only) and conditional (fixed and random effects) $R^2$ using the R~\citep{R} function `r.squaredGLMM' from the package \emph{MuMIn}~\citep{MuMIn}.

@@

            At low levels of disturbance, species' persistences increased with increasing proportions of the omnivory and direct competition motifs. Conversely, persistence decreased with as the  proportion of the apparent competition motif increased. Persistence was not significantly related to the proportion of the three-species chain (Figure~\ref{fig:prop_lmer_all}).

            At high levels of disturbance, persistence decreased as the proportion of the three-species chain increased, but increased as the proportion of all other motifs increased (Figure~\ref{fig:prop_lmer_all}).
            % This means that increased disturbance strengthened the effects of the three-species chain motif and changes direction of the effect from the apparent competition motif  (Table S4, \emph{Appendix SD}).
            The effect of omnivory on persistence was  positive but weaker at higher levels of disturbance, and the effect of direct competition was largely unaffected by disturbance level.

@@

            By fitting linear regressions of persistence against motif proportions for the species for each network separately, we can observe how the relationships between motifs and persistence change with the level of disturbance and global network structure. Network size had a much smaller effect than connectance (Fig. S3; \emph{Appendix SE}); we therefore focus on connectance here. 
            Networks with high connectance generally have more consistent relationships between persistence and motif participation, i.e., sharper peaks of the density distributions (Fig.~\ref{fig:density_prop_C}). 
            However, there was an interaction disturbance and connectance for the direct competition and omnivory motifs.

            At low and medium connectances there is little change in the proportion of positive slopes (\textgreater80\%) for the different levels of disturbance. 
            At high connectances, however, the proportion of positive slopes is lower overall and decreases with increasing disturbance. 
            This means that at low levels of connectance, an increased proportion of the direct competition motif increases persistence in most networks, while the opposite is true with high connectance. 
            The trend is reversed for the omnivory motif.

@@

        Motif participation varied with web size, connectance, and their interaction (Fig. S4, Table S6; \emph{Appendix SG}).
        Species in all networks appeared most frequently in the apparent competition motif, but the frequency of the apparent competition motif decreased with increasing network size and connectance.
        The frequency of the omnivory motif increased sharply with increasing connectance, while the frequency of the direct competition motif increased slightly with increasing network size. 
        Species' motif roles also vary with their in-degree and trophic level (Fig.~\ref{fig:motifs_vs_TL_and_deg}).
        Species with higher in-degree tend to have higher proportions of their motif role made up by omnivory. Proportions of the other three motifs tend to decline with increasing degree.

@@

        \begin{figure}
            \centering
            \includegraphics[width=.95\textwidth]{figures/persistence_positions_byprey.eps}
            \caption{A species' probability of persistence varied with the frequency of different motif positions in its role. Relationships are shown for the range 0-0.8 for all positions, but some positions (especially in the omnivory motif) are less common. Line colour indicates the motif containing a position. Line style indicates the number of prey within the motif for each position. Symbol shape indicates the number of predators within the motif for each position; note that only bottom-up extinctions were included in our simulations. The top set of lines corresponds to a probability of extinction for basal resources of 0.1. The middle set of lines corresponds to a probability of extinction for basal resources of 0.26, and the lowest set of lines corresponds to a probability of extinction for basal resources of 0.5.}
            \label{fig:persistence_positions_prey}
        \end{figure}

@@

        \subsubsection*{Degree and trophic level}

            The top position in the omnivory motif increased with increasing degree across all trophic levels.
            The frequency of the middle position increased with the interaction of degree and trophic level.
            The bottom position was generaly rare, especially for species with high degree or high trophic level (this suggests that the bottom position is frequently a basal resource).

            The frequency of the bottom position in the apparent competition motif decreased strongly with increasing degree, especially for species at a high trophic level.
            The frequency of the top position increased strongly with increasing degree and was not strongly related to trophic level.

            The frequency of the bottom position in the direct competition motif increased with increasing degree, especially for species at a high trophic level.
            The bottom position was very rare, especially for species with high degree (also suggesting that this position is usually a basal resource).

            The bottom position in the three-species chain was generally rare, especially for species with high degree or trophic level (again, this is likely to commonly be basal resources).
            The middle position in the three-species chain did not vary strongly with degree and declined slightly with increasing trophic level.
            The top position in the three-species chain increased with trophic level, degree, and their interaction.

        \subsubsection*{Network size and connectance}

            % Many of these are significant, but the slopes tend to be very shallow. TL and degree are more interesting - this is DEFINITELY SI.
            The frequency of the top position in the apparent competition motif did not vary strongly with network size or connectance.
            The frequency of the bottom position in the apparent competition motif decreased with increasing connectance but did not vary strongly vary with network size.

            The frequency of the top position in the direct competition motif increased slightly with the interaction of network size and connectance (i.e., in large and highly-connected networks).
            The frequency of the bottom position in the direct competition motif was low in all networks.

            The frequency of all three positions in the omnivory motif showed similar relationships with network size and connectance.
            Frequencies of these motifs increased with increasing connectance and were not strongly affected by network size.
            None of the positions in the three-species chain varied strongly with network size or connectance.

        A species probability of persisting is significantly related to the frequency of all positions, disturbance, and the interaction between frequency of each position and disturbance (Table S7, \emph{Appendix SH}).

        \begin{figure}
            \centering
            \includegraphics[width=.95\textwidth]{figures/persistence_positions_byprey.eps}
            \caption{A species' probability of persistence varied with the frequency of different motif positions in its role. Relationships are shown for the range 0-0.8 for all positions, but some positions (especially in the omnivory motif) are less common. Line colour indicates the motif containing a position. Line style indicates the number of prey within the motif for each position. Symbol shape indicates the number of predators within the motif for each position; note that only bottom-up extinctions were included in our simulations. The top set of lines corresponds to a probability of extinction for basal resources of 0.1. The middle set of lines corresponds to a probability of extinction for basal resources of 0.26, and the lowest set of lines corresponds to a probability of extinction for basal resources of 0.5.}
            \label{fig:persistence_positions_prey}
        \end{figure}

        The frequencies of motif positions in a species' role varied strongly and significantly with degree and trophic level, but only weakly with network size and connectance (\emph{Appendix SI}).

           \subsubsection*{Degree and trophic level}

            The top position in the omnivory motif increased with increasing degree across all trophic levels.
            The frequency of the middle position increased with the interaction of degree and trophic level.
            The bottom position was generaly rare, especially for species with high degree or high trophic level (this suggests that the bottom position is frequently a basal resource).

            The frequency of the bottom position in the apparent competition motif decreased strongly with increasing degree, especially for species at a high trophic level.
            The frequency of the top position increased strongly with increasing degree and was not strongly related to trophic level.

            The frequency of the bottom position in the direct competition motif increased with increasing degree, especially for species at a high trophic level.
            The bottom position was very rare, especially for species with high degree (also suggesting that this position is usually a basal resource).

            The bottom position in the three-species chain was generally rare, especially for species with high degree or trophic level (again, this is likely to commonly be basal resources).
            The middle position in the three-species chain did not vary strongly with degree and declined slightly with increasing trophic level.
            The top position in the three-species chain increased with trophic level, degree, and their interaction.

        \subsubsection*{Network size and connectance}

            % Many of these are significant, but the slopes tend to be very shallow. TL and degree are more interesting - this is DEFINITELY SI.
            The frequency of the top position in the apparent competition motif did not vary strongly with network size or connectance.
            The frequency of the bottom position in the apparent competition motif decreased with increasing connectance but did not vary strongly vary with network size.

            The frequency of the top position in the direct competition motif increased slightly with the interaction of network size and connectance (i.e., in large and highly-connected networks).
            The frequency of the bottom position in the direct competition motif was low in all networks.

            The frequency of all three positions in the omnivory motif showed similar relationships with network size and connectance.
            Frequencies of these motifs increased with increasing connectance and were not strongly affected by network size.
            None of the positions in the three-species chain varied strongly with network size or connectance.
